The title

### 2024-09-17 05_Manuscript/manuscript.Rmd
intro work

Intro paragraph - awareness of death, especially in light of COVID.

Various existential philosophers and psychologists have proposed ways in which we deal with the awareness of death and the anxiety this awareness often causes. Psychoanalyst Erik Erikson (1950) proposed that during mid-life one becomes acutely aware of their oncoming death and is motivated to care for things which will outlast themselves. He called this act of caring generativity. In *The Denial of Death*, philosopher Ernest Becker (1973) posits that humans undertake immortality projects to curb their sense of vulnerability to death. Similarly, psychiatrist Robert Jay Lifton (1979), a mentee of Erikson, described the awareness of death as being ever present and motivating us to create symbols, thereby allowing us to imagine ourselves as symbolically immortalized. Existential psychiatrist Irvin Yalom (2008) notes that many of his clients experiencing anxiety about their death take comfort in “rippling,” the idea that one’s lasting effects on the world will ripple out and influence the world after they have died.

Although these thinkers use different terminology, there are several common themes among their ideas. (1) Our physical death is an inevitability, and we often find our awareness of its inevitability to be aversive. This aversion may be referred to varyingly as angst, death-anxiety, despair, being-towards-death, terror, and so on. However, (2) we take comfort in the idea that other, non-physical parts of us continue to exist indefinitely after our biological death, through mechanisms such as the heroic archetype and symbolic self. (3) Finally, we can take action to promote these non-physical aspects of the self, such as through search for meaning, immortality project, care, generativity, and rippling.

One of these bodies of thought, called symbolic immortality, was originally theorized by Lifton (1979), who thought that awareness of death drives a fundamental human desire for a sense of continuity lasting beyond the lifespan. Essentially, humans are meaning-seeking creatures, and throughout our lives, this search for meaning involves an evolving psychological imagery of life and death. Death, or the transient nature of life, threatens our search of meaning. He thought that if we could achieve what we believe to be some form of immortality, we could overcome this loss of meaning, and the awareness of death could instead drive an inner vitality (imagery associated with connection, integrity, and movement). If this drive toward vitality is lost, we are vulnerable to what Lifton called a psychic numbness (imagery associated with separation, disintegration, and stasis). In Lifton's own words, "Death does indeed bring about biological and psychic annihilation. But life includes symbolic perceptions of connections that precede and outlast that annihilation" (1979, p. 18).

Lifton (1979) proposed five modes of experience or ways of achieving symbolic immortality: The biological (or biosocial) mode in which one lives on through their genetic and socio-cultural progeny, the creative mode in which one’s accomplishments and contribution outlast oneself, the natural mode in which one feels they are a part of the broader universe, the spiritual mode in which one seeks to transcend the physical realm to a higher spiritual realm beyond death, and the mode of experiential transcendence in which one experiences a phenomenological state of flow. The latter (experiential transcendence) must occur in the context of one of the other four to really be considered transcendent, but it is thought to have a great capacity to bring about personal change.

@@

TMT refers to a person's awareness of death as mortality salience (MS). The MS hypothesis of TMT posits than an increase in one's awareness of death causes an increase in compensatory behaviors to lower their death anxiety, either by distracting from the awareness of death or by the promotion of meaningful cultural worldviews. In the MS paradigm, experimentally priming a participant’s awareness of death (for example, by having participants write about death and then complete a distraction task) is thought to cause an increase in compensatory buffers. A meta-analysis of 277 experiments found mortality salience to have a robust, moderate overal effect size: *r*(276) = 0.35, *p* = .00 [@burke2010]. Altogether, these experiments provide convincing evidence for TMT and the MS hypothesis in particular.

@@

Drolet (1990) developed the Sense of Symbolic Immortality Scale based on Robert J. Lifton’s theory of symbolic immortality and its five modes of experience. Drolet studied 136 adults, ages 18-30 and 30-40, and hypothesized that those in their 30's (established adults) would have a greater sense of symbolic immortality than the young adults (18-30). The measure is inherently subjective, not only by the nature of self-report, but in that the scale seeks to measure what a person *believes* and how they *feel* about these subjects. The scale as a whole had a high internal consistency (alpha = .91) and test-retest reliability was r = .97. Internal consistency of subscales for the five theoretical modes of immortality was mixed. Of the five, spiritual immortality was the most distinct from the scale as a whole and the other subscales. (Note that this seems relevant to what we found in our own scale development. For more in the Discussion later on.) Factor analysis showed three factors, mapping onto biosocial, creative, and spiritual. The transcendent and natural items may be related to biosocial.

Moving beyond the scale development itself (still Drolet 1990), SSI correlated negatively with death anxiety (Templer's Death Anxiety Scale) and had a strong (r = .84) positive relationship with purpose in life (Maholick's Purpose in Life Test). In interpreting the very strong correlation, the author suggests that SSI is a broader construct than Purpose in Life and the scale itself may be less prone to social desirability effects than the PIL, although this had not been directly tested. Age group was also related, with established adults having a higher SSI, particularly in the biosocial and creative domains.

@@

Mathews and Mister (1987) also developed a scale involving symbolic immortality, sensation seeking, and psychic numbness. Some studies have used a revised version of the scale by Mathews and Kling (1988), who adapted it for a study on prosocial behavior in the context of nonprofit volunteer motivation. Mathews and Mister's study included 400 adults. Experiential transcendence was operationalized as something like Zuckerman's (1979) sensation seeking, which may not fully capture the original intent (the experience of losing oneself). Items were forced into five factors, and the five factors largely aligned with Lifton's constructs. Although internal consistency was at least acceptable for each factor, goodness of fit statistics are not reported. Mathews and Kling's (1988) study reported similar results for their revised scale. The items themselves seem to have more face validity than the scale by Drolet, but some factors seem more behavioral and unnecessarily specific: pertaining to one's religiosity or biological children, whereas Lifton's theory allows for a broader interpretation of these dimensions. The Nature and Creative factors seem most useful and theoretically aligned with Lifton.

>>>>

Death is something each of us must learn to cope with, whether in healthy ways or less so. These issues may be at front of mind for many in light of the COVID-19 pandemic. Various existential philosophers and psychologists have proposed ways in which we deal with the awareness of death and the anxiety this awareness often causes. Psychoanalyst Erik Erikson (1950) proposed that during mid-life one becomes acutely aware of their oncoming death and is motivated to care for things which will outlast themselves. He called this act of caring generativity. In *The Denial of Death*, philosopher Ernest Becker (1973) posits that humans undertake immortality projects to curb their sense of vulnerability to death. Similarly, psychiatrist Robert Jay Lifton (1979), a mentee of Erikson, described the awareness of death as being ever present and motivating us to create symbols, thereby allowing us to imagine ourselves as symbolically immortalized. Existential psychiatrist Irvin Yalom (2008) notes that many of his clients experiencing anxiety about their death take comfort in “rippling,” the idea that one’s lasting effects on the world will ripple out and influence the world after they have died.

Although these thinkers use different terminology, there are several common themes among their ideas. (1) Our physical death is an inevitability, and we often find our awareness of its inevitability to be aversive. This aversion may be referred to variously as angst, death-anxiety, despair, being-towards-death, terror, and so on. However, (2) we take comfort in the idea that other, non-physical parts of us continue to exist indefinitely after our biological death, through mechanisms such as the heroic archetype and symbolic self. (3) Finally, we can take action to promote these non-physical aspects of the self, such as through search for meaning, sense of immortality, care, generativity, and rippling.

One of these bodies of thought, called symbolic immortality, was originally theorized by Lifton (1979), who thought that awareness of death drives a fundamental human desire for a sense of continuity lasting beyond the lifespan. Essentially, humans are meaning-seeking creatures, and throughout our lives, this search for meaning involves an evolving psychological imagery of life and death. Death, or the transient nature of life, threatens our search for meaning. Lifton thought that if we could achieve what we believe to be some form of immortality, we could overcome this loss of meaning, and the awareness of death could instead drive an inner vitality (imagery associated with connection, integrity, and movement). If this drive toward vitality is lost, we are vulnerable to a psychic numbness or death-in-life (imagery associated with separation, disintegration, and stasis). In Lifton's own words, "Death does indeed bring about biological and psychic annihilation. But life includes symbolic perceptions of connections that precede and outlast that annihilation" (1979, p. 18).

Lifton (1979) proposed five modes of experience or ways of achieving symbolic immortality: The biological (or biosocial) mode in which one lives on through their genetic and sociocultural progeny, the creative mode in which one’s accomplishments and contribution outlast oneself, the natural mode in which one feels they are a part of the broader universe, the spiritual mode in which one seeks to transcend the physical realm to a higher spiritual realm beyond death, and the mode of experiential transcendence in which one experiences a phenomenological state of flow. The experiential mode must occur in the context of at least one of the other four to really be considered transcendent, but it is thought to have a great capacity to bring about personal change.

@@

TMT refers to a person's awareness of death as mortality salience (MS). The MS hypothesis of TMT posits than an increase in one's awareness of death causes an increase in compensatory behaviors to lower their death anxiety, either by distracting from the awareness of death or by the promotion of meaningful cultural worldviews. In the MS paradigm, experimentally priming a participant’s awareness of death (for example, by having participants write about death and then complete a distraction task) is thought to cause an increase in compensatory buffers. A meta-analysis of 277 experiments found mortality salience to have a robust, moderate overall effect size: *r*(276) = 0.35, *p* = .00 [@burke2010]. Altogether, these experiments provide convincing evidence for TMT and the MS hypothesis in particular.

@@

Drolet (1990) developed the Sense of Symbolic Immortality Scale based on Robert J. Lifton’s theory of symbolic immortality and its five modes of experience. Drolet studied 136 adults, ages 18-30 and 30-40, and hypothesized that those in their 30's (established adults) would have a greater sense of symbolic immortality than the young adults (18-30). The measure is inherently subjective, not only by the nature of self-report, but in that the scale seeks to measure what a person *believes* and how they *feel* about these subjects. The scale as a whole had a high internal consistency (*\$\\alpha\$* = .91) and test-retest reliability was *r* = .97. Internal consistency of subscales for the five theoretical modes of immortality was mixed. Of the five, spiritual immortality was the most distinct from the scale as a whole and the other subscales. Factor analysis showed three factors, mapping onto biosocial, creative, and spiritual. The transcendent and natural items may be closely related to biosocial.

Moving beyond the scale development itself (still Drolet 1990), SSI correlated negatively with death anxiety (Templer's Death Anxiety Scale) and had a strong (*r* = .84) positive relationship with purpose in life (Maholick's Purpose in Life Test). In interpreting the very strong correlation, the author suggests that SSI is a broader construct than Purpose in Life and the scale itself may be less prone to social desirability effects than the PIL, although this had not been directly tested. Age group was also related, with established adults having a higher SSI, particularly in the biosocial and creative domains.

@@

Mathews and Mister (1987) also developed a scale pertaining to symbolic immortality, sensation seeking, and psychic numbness in a study including 400 adults. Experiential transcendence was operationalized as something like Zuckerman's (1979) sensation seeking, which may not fully capture the original intent (the experience of losing oneself). Items were mapped onto five factors, and the five factors largely aligned with Lifton's constructs. Although internal consistency was at least acceptable for each factor, goodness of fit statistics are not reported. Some studies have used a revised version of the scale by Mathews and Kling (1988), who adapted it for a study on prosocial behavior in the context of nonprofit volunteer motivation. They reported similar results for their revised scale. The items on these scales seem to have more face validity than the scale by Drolet, but some factors seem more behavioral and unnecessarily specific: pertaining to one's religiosity or biological children, whereas Lifton's theory allows for a broader interpretation of these dimensions. The Nature and Creative factors seem most useful and theoretically aligned with Lifton.
